---
title: "Midterm Project"
author: "Joanna Meilhac"
date: "March 13, 2017"
output: html_document
---

I am preparing this data for analysis.
I will build a dataset from 3 different ones: osha.dbf, viol.dbf and accid.dbf.

I will clean each dataset and then join them together

# Part 1: Cleaning osha.dbf

## Reading the files

```{r message=FALSE}

library(magrittr)
library(plyr)
library(dplyr)
library(tidyr)
library(lubridate)
library(knitr)
library(foreign)
library(ggplot2)
a <- read.dbf("accid.dbf")
v <- read.dbf("viol.dbf")
o <- read.dbf("osha.dbf")


lookAcc <- read.dbf("lookups/acc.dbf")
lookhzs <- read.dbf("lookups/hzs.dbf")
lookocc <- read.dbf("lookups/occ.dbf")
lookSTD <- read.dbf("lookups/STD.dbf")
looksic <- read.dbf("lookups/sic.dbf")
lookNAIC <- read.dbf("lookups/naics.dbf")
lookscc <- read.dbf("lookups/scc.dbf")

```

## Cleaning up OSHA file



Some of the variables in the OSHA file:  

**CONTFLAG** : indicates whether record is continuation of previous record. This could be interesting if multiple establishments have a large number of proposed violations. A lot of the values in this column seem to be blank - meaning these violations would be the only ones for that company. We can see in the data that CONTFLAG has only two levels: 1 and 9. Let's count how many observations are in each level.

```{r}
count(filter(o, CONTFLAG == "1"))
count(filter(o, CONTFLAG == "9"))

```

There are only 2 observations which do not have a NA value in the CONTFLAG variable, which means only two establishments have a continuing record of violations. I will choose to take this variable out of the data.  
    
**JOBTITLE** : indicates the job classification of the officer conducting the inspection. I can take this variable out.  
  
**OWNERCODE**: Code if the type of owner = D (federal govt). When we count to see how many observations will have a value in this column, we find that 222 have codes for type of federal govt. I will take this variable out since it represents less than 0.3% of the data
```{r}
count(filter(o, OWNERCODE>0))
```

I do the same for other variables that are not of any interest to finding the most dangerous workplace. Here are the variables I will select:

OSHA1MOD - latest date activity applied against record  
PREVCTTYP - type of most recent OSHA activity, if any  
ACTIVITYNO - ID number for inspection record  
ESTABNAME - name of establishment  
SITEZIP - zipcode of site  
SITECITY - city code (see scc.dbf)  
SITECNTY - county code (see scc.dbf)  
OWNERTYPE - type of owner (private or government)  
OPENDATE - date of entry of inspection  
CLOSEDATE - date of exit from inspection site  
SIC - Standard Industrial Classification code  
NAICS - North American Industry Classification System  
INSPTYPE - inspection type  
UNION - employees represented by union or not  
LWDIRATE - Bureau of Labor Statistics last workday injury rate  
PALITIGN - time spent preparing for court and testifying  
PASUMHOURS - sum of all inspection activity hours reported  
PENREMIT - total remitted penalty dollars  
TOTPENLTY - total current penalties issued  
TOTALVIOLS - total number of violations issued  
TOTSERIOUS - total number of serious, willful, and repeat violations issued for inspection  

  
```{r}

o <- select(o, OSHA1MOD, PREVCTTYP, ACTIVITYNO, ESTABNAME, SITEZIP, SITECITY,  SITECNTY, OWNERTYPE, OPENDATE, CLOSEDATE, SIC, NAICS, INSPTYPE, UNION, LWDIRATE, PALITIGN, PASUMHOURS, PENREMIT, TOTPENLTY, TOTALVIOLS, TOTSERIOUS)
```
  
**1. Change coding of PREVCTTYP**
```{r}
levels(o$PREVCTTYP)
levels(o$PREVCTTYP) <- c("accident", "complaint", "inspection", "referral")
```
  
**2. Get SIC data from lookups**
```{r warnring = FALSE}
o <- left_join(o, looksic, by = "SIC")
o <- select(o, -SIC)
```
  
**3. NAIC**
```{r warning = FALSE}
o <- left_join(o, lookNAIC, by = "NAICS")
o <- select(o, -NAICS)
# If I sum all the NA values in the NAICS column and compare it to the NA values in the SIC column, we can see that the SIC column has a value for every observation, whereas NAICSTEXT column has 74645 NA values. For this reason, I will decide to keep the Standard Industry Classification and get rid of the North American Industry Classification.
sum(is.na(o$INDUSTRY)==TRUE)
sum(is.na(o$NAICSTEXT)==TRUE)
o <- select(o, -NAICSTEXT)
```
  
**4. Change coding of OWNERTYPE**
```{r}
levels(o$OWNERTYPE)
levels(o$OWNERTYPE) <- c("private", "local government", "state government", "federal government")
```
  
**5. SITECITY**
```{r warning = FALSE}
# Looking at lookscc, the coding for the different towns are made up of both the county code and the city code. I will make a new variable to identify this code.
lookscc <- lookscc[(lookscc$STATE == "MA"),]
lookscc$CITYCODE <- with(lookscc, interaction(COUNTY, CITY))
lookscc <- select(lookscc, CITYCODE, NAME)
o$CITYCODE <- with(o, interaction(SITECNTY, SITECITY))
o <- left_join(o, lookscc, by = "CITYCODE")
o <- select(o, -CITYCODE, -SITECNTY, -SITECITY)
colnames(o)[19] <- "CITY"
```
  
**6. Change coding of INSPTYPE**
```{r}
levels(o$INSPTYPE)
# If I look in osha.txt, it shows me a differentiation between programmed and unprogrammed inspections. I will split this variable into 2 different variables; the type of programmed inspection, and the type of unprogrammed inspection.

programmed <- mutate(o, PROG_INSPTYPE = INSPTYPE)%>%
  select(PROG_INSPTYPE)
levels(programmed$PROG_INSPTYPE) <- c(NA, NA, NA, NA, NA, NA, NA, "planned", "related", NA, "other", "non-inspection")

unprogrammed <- mutate(o, UNPROG_INSPTYPE = INSPTYPE)%>%
  select(UNPROG_INSPTYPE)
levels(unprogrammed$UNPROG_INSPTYPE) <- c("fatality/catastrophe", "complaint", "referral", "monitoring", "variance", "follow-up", "related", NA, NA, "other", NA, NA)


o <- o%>%
  bind_cols(programmed)%>%
  bind_cols(unprogrammed)%>%
  select(-INSPTYPE)

```
  
**7. Convert OSHA1MOD from integer to date**
```{r warning = FALSE}
o$OSHA1MOD <- ymd(o$OSHA1MOD)
```
  
**8. OPENDATE**
```{r warning = FALSE}
o$OPENDATE <- ymd(o$OPENDATE)
```
  
**9. CLOSEDATE**
```{r warning = FALSE}
o$CLOSEDATE <- ymd(o$CLOSEDATE)
```
  
**10. Change the value of 0 in PASUMHOURS to NA since it is impossible to spend no time on the inspection if there was an inspection**
```{r}
is.na(o$PASUMHOURS) <- !o$PASUMHOURS
```
  
**11. Change value of 00000 in SITEZIP to NA**
```{r}
o$SITEZIP[o$SITEZIP == "00000"] <-  NA
```

**12. Create DURATION variable that gives the number of days of inspection, for the inspections that lasted longer than one day**
```{r}
o <- mutate(o, INTERVAL=OPENDATE%--%CLOSEDATE)%>%
  mutate(DURATION1 = as.duration(INTERVAL))%>%
  mutate(DURATION = as.numeric(DURATION1, "days"))%>%
  select(-INTERVAL, -DURATION1)

is.na(o$DURATION) <- !o$DURATION
```

```{r echo = FALSE}
qplot(x=DURATION, y=TOTSERIOUS, data=o,geom="point", color=OWNERTYPE)+xlim(c(0, 100))
```
  
  
  
This histogram shows us that most of the inspections last less than 5 days:
```{r echo = FALSE, message = FALSE, warning = FALSE}
ggplot(o) + geom_histogram(aes(x=DURATION), fill="red", alpha = .4) + xlim(c(0, 150))

```

  
Save the cleaned data:
```{r}
o <- distinct(o)
write.dbf(o, file="osha1.dbf")
```
```{r echo = FALSE}
kable(o[1:10, ], caption = "A table of the cleaned osha.dbf data")
```



# Part Two: Cleaning viol.dbf

Looking at all the variables for relevant ones: I will keep the following variables:

 EMPHASIS - Value = X if  case is egregious -- I will have to change the coding for this variable  
 GRAVITY - indicates level of potential harm, on a scale of 0 - 10  
 PENCURRENT - current penalty OSHA has collected or intends to collect  
 VIOLTYPE - current type of violation -- change the coding of values  
 STD - kind of violation (federal or state standard cited)  
 INSTANCES - number of instances of violation of standard related event  
 REC - related event code -- I will use this to create a FATALITIES variable  
 NUMEXPOSED - number of employees exposed to hazard  
 DISPEVT - disposition of contested violation -- make a variable for court decisions  
 HAZCAT - Hazardous category  


```{r}
v <- select(v, ACTIVITYNO, EMPHASIS, GRAVITY, PENCURRENT, INSTANCES, VIOLTYPE, STD, REC, NUMEXPOSED, DISPEVT, HAZCAT)
```
  
**1. Change coding of Emphasis:**
```{r}
levels(v$EMPHASIS)
levels(v$EMPHASIS) <- c("egregious")
```
  
**2. Change coding of STD**
```{r}
colnames(lookSTD) <- c("STATE", "STD", "TEXT")
v <- left_join(v, lookSTD, by="STD") 

v <- select(v, -STATE, -STD)
colnames(v)[11] <- "STD"
```
  
**3. Change coding of VIOLTYPE**
```{r}
levels(v$VIOLTYPE)
levels(v$VIOLTYPE) <- c("other", "repeat", "serious", "unclassified", "willful")
```
  
**4. Change REC so that it only codes for fatalities**
```{r}
levels(v$REC)
# Since the text file for VIOL does not give us information for some of the values in the data, I will only use the value assigned to fatality/catastrophe, as I think this represents the more dangerous workplaces.
levels(v$REC) <- c(NA, "Fatality/Catastrophe", NA, NA, NA, NA ,NA, NA, NA, NA)
colnames(v)[7] <- "FATALITY_CATASTROPHE"
```


  
**5. Change DISPEVT so that it only codes for court cases**
```{r}
levels(v$DISPEVT)
levels(v$DISPEVT) <- c("appeals court decision", "supreme court decision", "government dismissed case", "formal settlement agreement", "administrative law judge decision", NA, NA, NA)

```
  
**6. Change coding of GRAVITY**
```{r}
levels(v$GRAVITY)
levels(v$GRAVITY) <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```
  
And finally, we save the file:
```{r}
v <- distinct(v)
write.dbf(v, file="violations.dbf")
```

```{r echo = FALSE}
kable(v[1:10, ], caption = "A table of the cleaned viol.dbf data")

```


# Part Three: Cleaning accid.dbf

Looking at all the variables for relevant ones: I will keep all the variables except for SITESTATE, NAME, and RELINSP.

```{r}

if(sum(a$SITESTATE=="MA")==dim(a)[1]){a%<>% select(-SITESTATE)}

a <- select(a, -NAME, -RELINSP)
```
  
**1. Change factors of DEGREE**
```{r}
# 1 - Change factors of DEGREE
levels(a$DEGREE)
levels(a$DEGREE) <- c("no injury", "fatality", "hospitalized", "non-hospitalized")
```
  
**2. NATURE of injury**
```{r}
levels(lookAcc$CATEGORY)
sum(lookAcc$CATEGORY=="NATUR-INJ")
levels(a$NATURE)

nature <- lookAcc[(lookAcc$CATEGORY == "NATUR-INJ"),]
dim(nature)
# Take CATEGORY out of new dataframe and change column names to join with accid.dbf
nature <- select(nature, CODE, VALUE)
colnames(nature) <- c("NATURE", "VALUE")
a <- left_join(a, nature, by = "NATURE")

# take out the column with the coding and rename the column with the values
a <- select(a, -NATURE)
colnames(a)[13] <- "NATURE"
```

**3. BODYPART**
```{r}
sum(lookAcc$CATEGORY=="PART-BODY")
levels(a$BODYPART)
parts <- lookAcc[(lookAcc$CATEGORY == "PART-BODY"),]

parts <- select(parts, CODE, VALUE)
colnames(parts) <- c("BODYPART", "VALUE")
a <- left_join(a, parts, by= "BODYPART")

a <- select(a, -BODYPART)
colnames(a)[13] <- "BODYPART"
```
  
**4. SOURCE of injury**
```{r}
sum(lookAcc$CATEGORY == "SOURC-INJ")
levels(a$SOURCE)
source <- lookAcc[(lookAcc$CATEGORY == "SOURC-INJ"),]

source <- select(source, CODE, VALUE)
colnames(source) <- c("SOURCE", "VALUE")
a <- left_join(a, source, by = "SOURCE")

a <- select(a, -SOURCE)
colnames(a)[13] <- "SOURCE"
```
  
**5. EVENT type**
```{r}
sum(lookAcc$CATEGORY == "EVENT-TYP")
levels(a$EVENT)
events <- lookAcc[(lookAcc$CATEGORY == "EVENT-TYP"),]

events <- select(events, CODE, VALUE)
colnames(events) <- c("EVENT", "VALUE")
a <- left_join(a, events, by = "EVENT")

a <- select(a, -EVENT)
colnames(a)[13] <- "EVENT"
```
  
**6. ENVIRON - contributing factor**
```{r}
sum(lookAcc$CATEGORY == "ENVIR-FAC")
levels(a$ENVIRON)
environ <- lookAcc[(lookAcc$CATEGORY == "ENVIR-FAC"),]

environ <- select(environ, CODE, VALUE)
colnames(environ) <- c("ENVIRON", "VALUE")
a <- left_join(a, environ, by = "ENVIRON")

a <- select(a, -ENVIRON)
colnames(a)[13] <- "ENVIRON"
```
  
**7. HUMAN factor**
```{r}
sum(lookAcc$CATEGORY == "HUMAN-FAC")
levels(a$HUMAN)
humanfac <- lookAcc[(lookAcc$CATEGORY == "HUMAN-FAC"),]

humanfac <- select(humanfac, CODE, VALUE)
colnames(humanfac) <- c("HUMAN", "VALUE")
a <- left_join(a, humanfac, by = "HUMAN")

a <- select(a, -HUMAN)
colnames(a)[13] <- "HUMAN"
```
  
**8. Change coding of TASK**
```{r}
levels(a$TASK)
levels(a$TASK) <- c(NA, "regularly assigned task", "task other than regularly assigned")
```
  
**9. HAZSUB**  
```{r}
colnames(lookhzs) <- c("HAZSUB", "VALUE")
a <- left_join(a, lookhzs, by = "HAZSUB")

a <- select(a, -HAZSUB)
colnames(a)[13] <- "HAZSUB"
```
  
**10. OCCUPATION**
```{r}
colnames(lookocc) <- c("OCC_CODE", "OCCUPATION")
a <- left_join(a, lookocc, by = "OCC_CODE")
a <- select(a, -OCC_CODE)
```
  
**11. Change age of 0 to NA**
```{r}
is.na(a$AGE) <- !a$AGE

```
  
And finally, save the file:
```{r}
a <- distinct(a)
write.dbf(a, file = "accidents.dbf")
```


```{r, echo = FALSE}

kable(a[1:10, ], caption = "A table of the cleaned accid.dbf")
```
  
  
  
# Part 4: Joining all the datasets together

```{r}

finaldata <- left_join(o, v, by = "ACTIVITYNO")
finaldata <- left_join(finaldata, a, by = "ACTIVITYNO")


write.dbf(finaldata, file = "cleandata.dbf")
```


```{r echo = FALSE}

kable(finaldata[1:10, ], caption = "A table of the cleaned data")
```
  
  
This graph shows us two things: before 1984, there was no coding for GRAVITY. Secondly, the values for GRAVITY are more extreme after about 1992, since there are no establishments that have a GRAVITY value of 7, 8, or 9.
```{r echo = FALSE}
ggplot(finaldata)+ geom_point(aes(x=OPENDATE, y=GRAVITY), color="turquoise", size=2)

set=subset(finaldata, subset = VIOLTYPE == "serious")

```







